\vspace{-2.0cm}The effect of nitrogen application time on infection and development of the Wheat Streak Mosaic Virus (WSMV)

### 2015-10-22 leslieandkenny.Rnw
finished my sections, fixed typos and wording, I think this draft is ready to go

\noindent Nar B. Ranabhat is a PhD candidate in the Department of Land Resources and Environmental Science. He is currently working on several projects regarding the spread and development of the Wheat Streak Mosaic Virus (WSMV) in Winter Wheat. Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently sitting in on Mixed Models. Nar has asked for our help in building a model, fitting it in R, and interpreting the results.

@@

We address the following questions in this report. The first three questions are related to model selection, and the last question is related to interpretation after a model is selected.
\begin{itemize}
\item Is there evidence that the interaction between nitrogen application and variety differs between inoculated and control plots?
\item Is there evidence that the nitrogen application effect varies across varieties? Is there evidence that the nitrogen application effect differs between inoculated and control?
\item Is there evidence that the variety effect differs between inoculated and control?
\item Conditional on the results to the above questions, how do we appropriately describe the estimated effects for variety, nitrogen application time, and inoculation on the probability of virus infection?
\end{itemize}

@@

\noindent One serious problem that can occur when working with counts is
the issue of separation. This is when one variable or a combination of factors
is perfectly associated with the response. For example, if a certain
treatment results in an infection count of zero, then knowing that a leaf
recieved that treatment determines that the leaf will not get infected,
regardless of other variables. This causes difficulty in estimating the
effect of that treatment on the infection rate, and results a large standard
error because the effect cannot be precisely determined. \\

@@

\begin{figure}[h!]
\begin{center}
\caption{Illustration of separation.}
<<separation,echo=FALSE>>=
par(mfrow = c(1, 2), mar = c(4, 4, 2, 2))
curve(exp(x)/(1+exp(x)), from = -10, to = 10, ylim = c(0, 1),
      xlab = expression(logit(p[i])), ylab = expression(p[i]))
segments(x0 = -10, x1 = log(1/59), y0 = 1/60, lty = 2, lwd = 2)
segments(x0 = log(1/59), y0 = 0, y1 = 1/60, lty = 2, lwd = 2)

@@

\noindent The following sections present examples of issues that could
arise while analyzing Nar's data, and our recommendations on dealing with
them. The examples all use simulated datasets.

@@

\begin{figure}[h!]
\caption{Example interaction plot using \texttt{qplot}.}
<<interactionplot,echo=FALSE,message=FALSE,out.height='0.3\\linewidth',fig.height=3>>=
require(ggplot2)
qplot(x = n.trt, y = infected, geom = c("boxplot", "point"), color = status,
      facets = .~variety, data = fert.sim1) +
  theme(axis.text.x = element_text(angle = 90))
@
\end{figure}

\noindent A graphical analysis is also a good way to see if separation
is present in the data. In Nar's dataset, 67 of the 180 counts are
0. This is a large fraction of the data, so it is very important to look
for possible separation. We recommend sorting the data and creating a
heatmap to look for patterns of zeros. Figure 2 shows an example using
simulated data. The most serious problems will come from the zeros, so we
chose a contrasting color (black) for plots with infection counts of 0. \\

\begin{figure}[h!]
\caption{Example heatmap visualization for identifying patterns of zeros.}
<<heatmap,echo=FALSE,message=FALSE,out.height='0.3\\linewidth',fig.height=3>>=
## Heatmap with meaningful ordering

@@

\noindent If a treatment has many zeros across all varieties or a variety
has many zeros, Nar should consider whether he has reason to believe
that treatment or variety has qualitative differences from the others
that would cause it to have a much lower infection rate. These data would
be considered as coming from a different population than the data with
fewer zeros. \emph{If and only if it is justified}, Nar could omit all
of the data for this treatment or variety when fitting the model. The scope
of inference for this model must then be limited to varieties and
treatments that have a nonzero infection rate. \\

@@

\subsection{Binomial Generalized Linear Model}

\noindent Because the response variable is a Binomial count, and because
the design includes random assignment of variety to rows within blocks,
a Binomial generalized linear model is appropriate. We recommend using
a three-way interaction to estimate the probability of infection for each
variety/treatment/inoculation status combination and a block/variety
interaction to control for differences between blocks and between rows
in the same block.\\

@@

@@

@@

@@

\noindent Since the inoculated plants were planted in the center of the
plots, we expect to see spatial dependence where plants near the centers
are more likely to become infected than plants further from the center.
Then individual plants could have infection probabilities that differ from
the infection rate estimated for the plot they are in. This would result
in the data could having more variation than what the Binomial model
accounts for, a situation known as overdispersion. We observed
overdispersion in Nar's data, and there are no simple tools
for including overdispersion in mixed models. Therefore, the standard
errors in the mixed model output would be based on an incorrect estimate
of the variance and might not be trustworthy. The standard errors in the GLMM model would be smaller than they should be, causing the p-values for effects of interest to be artificially small. This problem is easily remedied by using a quasibinomial fixed-effects GLM model. \\

\noindent We illustrate this problem in the output below. We fit a quasibinomial logistic regression model, and a binomial generalized linear mixed effects model to the real data sent to us by Nar (details on model fitting in section \ref{model} and \ref{glmmfit}). The coefficients, standard errors, and tests are shown for the three way interaction coefficients in each model. The estimates are similar between the two models, but the standard errors are much larger in the quasibinomial GLM model. Without accounting for overdispersion, the standard errors for the estimates in the mixed effects model are artificially small. For this reason, we advise Nar not to trust the results from the mixed effects model.

@@

\noindent Rows are nested within blocks in the study design, and each row is randomly assigned to a variety. The design's nesting structure can be accounted for in a GLMM model  by including random effects for row and block. Originally, however, random effects were developed to reflect a feature of the study design. For example, if $10$ operators were chosen at random from a larger population of operators, {\it operator} would be modeled as a random effect. In this case, the sample of operators truly is random in the design, and the operators selected are intended to represent a larger population of operators. In Nar's study design, row and block are not truly random effects because they were not selected from a larger population of rows and blocks. It is important to recognize this and understand that the only reason row and block are included as random effects in the GLMM model is to account for the nesting structure of row within block. As a result, we think it is more appropriate to model the nested design using fixed effects in the binomial generalized linear model, and this strategy is more consistent with the study design. \\

@@

\subsection{Fitting the GLM}
\label{model}

@@

\noindent If it is decided that certain observations come from an extremely
resistant population and can be omitted from the current analysis:

@@

@@

In this section, we describe how to go through the process of model selection. As discussed in the previous section, we recommend starting with the most complicated three-way interaction model and then assessing which terms can be removed. In general, start by considering removal of the highest order terms. \\

\noindent We'll start by assessing evidence for the three-way interaction term. One way to consider it's removal would be to fit a model without the three-way interaction, and then conduct a drop-in-deviance F-test comparing the full and reduced models, as you learned to do in Stat $512$. This method can be time consuming, so instead we recommend using the \verb+Anova+ function in the \verb+car+ package with the option \verb+test="F"+. This provides a {\bf Type II} sums of squares table, and the tests provided are similar to extra sums of squares F-tests based on the Pearson residuals.
%(provide reference here to info about pearson resids vs deviance resids)
The tests in this table are conditional upon terms of the same and lower order being in the model. For example, the \verb+n.trt:status2+ row in the table tests for removal of the $n.trt*status2$ two-way interaction, conditional upon all other two way interactions and main effects being in the model. These tables can be used to assess all the effects at the same level together. If you choose to remove a term, however, it is important to fit a new model and reconsider the Anova table every time a simpler model is selected. This is necessary because the estimate of overdispersion changes when the model changes, causing the tests in the table to change as well. \\

@@

\section{References}
%\noindent Ramsey, F.L., Schafer, D.W. (2013). {\it The Statistical Sleuth: A Course in Methods of Data Analysis, Third Edition}. Boston, MA: Brooks/Cole, Cengage Learning.\\

@@

\noindent Sloderbeck, J., Michaud, P., Whitworth, Robert. "Wheat Pests." CurlMite. Kansas State University, 1 May 2008. Web. 18 Sept. 2015.\\
{\footnotesize\verb|http://entomology.k-state.edu/extension/insect-information/crop-pests/wheat/curlmite.html|.}

>>>>

\noindent Nar B. Ranabhat is a PhD candidate in the Department of Land Resources and Environmental Science. He is currently working on several projects regarding the spread and development of the Wheat Streak Mosaic Virus (WSMV) in Winter Wheat. Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently sitting in on Mixed Models. Nar has asked for our help in building a model, fitting it in R, and interpreting the results. In this report we describe the study and its potential issues, present our recommendations, and provide example R code.

@@

\noindent We address the following questions in this report. The first three questions are related to model selection, and the last question is related to interpretation after a model is selected.
\begin{itemize}
\item Is there evidence that the interaction between nitrogen application and variety differs between inoculated and control plots?
\item Is there evidence that the nitrogen application effect varies across varieties? Is there evidence that the nitrogen application effect differs between inoculated and control?
\item Is there evidence that the variety effect differs between inoculated and control?
\item Conditional on the results to the above questions, how do we appropriately describe the estimated effects for variety, nitrogen application time, and inoculation on the probability of virus infection?
\end{itemize}

@@

\noindent One serious problem that can occur when working with counts is
separation. This is when one variable or a combination of factors
is perfectly associated with the response. For example, if a certain
treatment results in an infection count of zero, then knowing that a leaf
received that treatment determines that the leaf will not get infected,
regardless of other variables. This causes difficulty in estimating the
effect of that treatment on the infection rate, and results in a large standard
error because the effect cannot be precisely determined. \\

@@

\begin{figure}[h!]
\label{sep}
\begin{center}
\caption{Illustration of separation.}
<<separation,echo=FALSE,out.height='0.3\\linewidth',fig.height=3>>=
par(mfrow = c(1, 2), mar = c(4, 4, 2, 2))
curve(exp(x)/(1+exp(x)), from = -10, to = 10, ylim = c(0, 1),
      xlab = expression(logit(p[i])), ylab = expression(p[i]))
segments(x0 = -10, x1 = log(1/59), y0 = 1/60, lty = 2, lwd = 2)
segments(x0 = log(1/59), y0 = 0, y1 = 1/60, lty = 2, lwd = 2)

@@

\noindent The following sections present examples of issues that could
arise while analyzing Nar's data, and our recommendations on dealing with
them. Except where stated otherwise, the examples use simulated datasets.

@@

\begin{figure}[h!]
\label{qp}
\caption{Example interaction plot using \texttt{qplot}.}
<<interactionplot,echo=FALSE,message=FALSE,out.height='0.3\\linewidth',fig.height=3>>=
require(ggplot2)
qplot(x = n.trt, y = infected, geom = c("boxplot", "point"), color = status,
      facets = .~variety, data = fert.sim1) +
  theme(axis.text.x = element_text(angle = 90))
@
\end{figure}

\noindent A graphical analysis is also a good way to see if separation
is present in the data. In Nar's dataset, 67 of the 180 counts are
0. This is a large fraction of the data, so it is very important to identify
possible separation. We recommend sorting the data and creating a
heatmap to look for patterns of zeros. Figure 3 shows an example using
simulated data. The most serious problems will come from the zeros, so we
chose a contrasting color (black) for plots with infection counts of 0.
The R code to produce this graphic appears in the appendix. \\

\begin{figure}[h!]
\label{hm}
\caption{Example heatmap visualization for identifying patterns of zeros.}
<<heatmap,echo=FALSE,message=FALSE,out.height='0.3\\linewidth',fig.height=3>>=
## Heatmap with meaningful ordering

@@

\noindent If a treatment has many zeros across all varieties or a variety
has many zeros, Nar should consider whether he has reason to believe
that treatment or variety has qualitative differences from the others
that would cause it to have a much lower infection rate. These data would
be considered as coming from a different population than the data with
fewer zeros. \textbf{If and only if it is justified}, Nar could omit all
of the data for this treatment or variety when fitting the model. The scope
of inference for this model must then be limited to varieties and
treatments that have a nonzero infection rate. \\

@@

\subsection{Quasibinomial Generalized Linear Model}

\noindent We recommend starting with a quasibinomial generalized linear
model. This model includes a three-way interaction to estimate the
infection probability for each variety/treatment/inoculation status
combination, and a block/variety interaction to control for differences
between blocks and between rows in the same block.\\

@@

\noindent The overdispersion parameter accounts for slight violations
of these assumptions. We expect the plants within a single plot to
have different infection probabilities based on their proximity to
plants where mites are present. This issue manifests itself in the form
of excess variation because the probability that a particular plant
gets infected will not be the same as the estimated probability for
plants in its plot. The overdispersion is an estimate of this extra
variability; without it the standard errors will be underestimated. \\

@@

\noindent Another option is to use random effects to model the nested
structure of row within block. This model includes a three-way
interaction of fixed effects and does not include an overdispersion
parameter. In mixed models, the random effects are meant to explain
all the variation in the data, so overdispersed models are not
used in this setting. \\

@@

\noindent The assumptions for this model are that plants within a plot
are independent and have the same probability of being infected, that
blocks are independent, that rows in different blocks are independent, that
plots in different rows are independent given the other variables in the
model, and that there are no additional sources of variation besides
block, row, and the Binomial distribution for the infection counts. \\

\noindent This model accounts for block-to-block and row-to-row variation
by including variance terms for the block and row random effects. The
benefit of this is that the amount of variation described by the blocks
and rows will be estimated. The drawback is that, if there is additional
variation not explained by these random effects, the model will fit poorly.

@@

\noindent Since the inoculated plants were planted in the center of the
plots, we expect to see spatial dependence where plants near the centers
are more likely to become infected than plants further from the center.
Then individual plants could have infection probabilities that differ from
the infection rate estimated for the plot they are in. This would result
in the data having more variation than what the Binomial model accounts for,
a situation known as overdispersion. We observed overdispersion in Nar's
data, and mixed models have limited ability to account for overdispersion
beyond what can be explained by the variables in the model. Therefore,
the standard errors in the mixed model output would be based on an
incorrect estimate of the variance and might not be trustworthy. The standard errors in the GLMM model would be smaller than they should be, causing the p-values for effects of interest to be artificially small. This problem is easily remedied by using a quasibinomial fixed-effects GLM model. \\

\noindent We illustrate this problem in the output below. We fit a quasibinomial logistic regression model, and a binomial generalized linear mixed effects model to the real data sent to us by Nar (details on model fitting in section \ref{model} and \ref{glmmfit}). The coefficients, standard errors, and tests are shown for the three way interaction coefficients in each model. The estimates are similar between the two models, but the standard errors are much larger in the quasibinomial GLM model. Without accounting for overdispersion, the standard errors for the estimates in the mixed effects model are too small. For this reason, we advise Nar not to trust the results from the mixed effects model.

@@

\noindent Rows are nested within blocks in the study design, and each row is randomly assigned to a variety. The nesting structure of the design can be accounted for in a GLMM model  by including random effects for row and block. Originally, however, random effects were developed to reflect a feature of the study design. For example, if $10$ operators were chosen at random from a larger population of operators, {\it operator} would be modeled as a random effect. In this case, the sample of operators truly is random in the design, and the operators selected are intended to represent a larger population of operators. In Nar's study design, row and block are not truly random effects because they were not selected from a larger population of rows and blocks. It is important to recognize this and understand that the only reason row and block are included as random effects in the GLMM model is to account for the nesting structure of row within block. As a result, we think it is more appropriate to model the nested design using fixed effects in the binomial generalized linear model, and this strategy is more consistent with the study design. \\

@@

\pagebreak
\subsection{Fitting the GLM}
\label{model}

@@

\noindent The summary output should be examined for unexpectedly large
estimates or standard errors. A warning message may or may not appear in
this situation. If it is decided that certain observations come from an
extremely resistant population and can be omitted from the current analysis:

@@

To fit the three-way interaction model:

<<glmm1,eval=FALSE>>=
glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/variety),
                  control = glmerControl(optimizer = "bobyqa"),
                  family = binomial, data = fert.sim1)
@

\noindent The \verb|control = glmerControl(optimizer = "bobyqa")| argument
tells the \verb|glmer| function which numerical optimizer to use to find
the estimates. Several optimizers are included with R, but \verb|bobyqa|
works well on problems that are mathematically difficult. We recommend using
it for all \verb|glmer| fits. \\

\noindent The following warning message will probably appear:

<<glmmhide1,echo=FALSE,cache=TRUE>>=
glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/variety),
                  control = glmerControl(optimizer = "bobyqa"),
                  family = binomial, data = fert.sim1)
@

\noindent This message alone does not indicate any problems, but it
appears because the model has a large number of parameters, which could
cause the numerical optimizer to take many iterations to converge. By
default, the algorithm will run for a maximum of 10,000 iterations. That
number can be increased by specifying a larger \verb|maxfun| option:

<<glmm2,eval=FALSE>>=
glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/variety),
                  control = glmerControl(optimizer = "bobyqa",
                                         optCtrl = list("maxfun" = 20000)),
                  family = binomial, data = fert.sim1)
@

\noindent If a warning message says something like ``maximum number of
function evaluations exceeded'' or mentions ``max$\lvert$grad$\rvert$''
and a tolerance level, then the optimizer ran for the full number of
iterations without finding a solution. Increasing \verb|maxfun| as above
may solve the problem. \\

\noindent If messages about eigenvalues or ``unable to evaluate scaled
gradient'' occur:

<<glmmhide2,echo=FALSE,cache=TRUE>>=
glmmPlot <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|plot),
                  control = glmerControl(optimizer = "bobyqa",
                                         optCtrl = list("maxfun" = 20000)),
                  family = binomial, data = fert2)
glmmVar <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
                   (1|block/variety),
                 control = glmerControl(optimizer = "bobyqa",
                                        optCtrl = list("maxfun" = 20000)),
                 family = binomial, data = fert2)
@

\noindent These indicate that unique coefficient estimates do not exist.
This is most likely caused by separation, so the output should be
examined for large standard errors. The best solution would be to fit
a simpler model (such as one with fewer interaction terms) or to remove
problematic variables or data. The decision to remove any terms or data
must be justified by prior knowledge and not based on the R output.
If this cannot be justified, then the generalized linear mixed model should
not be used. \\

@@

In this section, we describe how to go through the process of model selection. As discussed in the previous sections, we recommend starting with the most complicated three-way interaction model and then assessing which terms can be removed. In general, start by considering removal of the highest order terms. \\

\noindent We'll start by assessing evidence for the three-way interaction term. One way to consider its removal would be to fit a model without the three-way interaction, and then conduct a drop-in-deviance F-test comparing the full and reduced models, as you learned to do in Stat $512$. This method can be time consuming, so instead we recommend using the \verb+Anova+ function in the \verb+car+ package with the option \verb+test="F"+. This provides a {\bf Type II} sums of squares table, and the tests provided are similar to extra sums of squares F-tests based on the Pearson residuals.
%(provide reference here to info about pearson resids vs deviance resids)
The tests in this table are conditional upon terms of the same and lower order being in the model. For example, the \verb+n.trt:status2+ row in the table tests for removal of the $n.trt*status2$ two-way interaction, conditional upon all other two way interactions and main effects being in the model. These tables can be used to assess all the effects at the same level together. If you choose to remove a term, however, it is important to fit a new model and reconsider the Anova table every time a simpler model is selected. This is necessary because the estimate of overdispersion changes when the model changes, causing the tests in the table to change as well. \\

@@

\pagebreak
\section{References}
%\noindent Ramsey, F.L., Schafer, D.W. (2013). {\it The Statistical Sleuth: A Course in Methods of Data Analysis, Third Edition}. Boston, MA: Brooks/Cole, Cengage Learning.\\

@@

\noindent Sloderbeck, J., Michaud, P., Whitworth, Robert. ``Wheat Pests.'' CurlMite. Kansas State University, 1 May 2008. Web. 18 Sept. 2015.\\
{\footnotesize\verb|http://entomology.k-state.edu/extension/insect-information/crop-pests/wheat/curlmite.html|.}
